Keystoneness, centrality, and the structural controllability of ecological networks

### 2019-01-13 paper/drafts_and_revisions/2019-01-14/reviewers-response.tex

\documentclass[a4paper]{artikel1}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Keystoneness, centrality, and the structural controllability of ecological networks},
            pdfauthor={Fernando Cagua, Kate L. Wootton, Daniel B. Stouffer},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Keystoneness, centrality, and the structural controllability of
ecological networks}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
\subtitle{JEcol-2018-0545 - Detailed response to comments}
  \author{Fernando Cagua, Kate L. Wootton, Daniel B. Stouffer}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

\usepackage{setspace}
\usepackage{xr}
\usepackage{lineno}
\newcommand{\R}[1]{\label{#1}\linelabel{#1}}
\newcommand{\lr}[1]{page~\pageref{#1}, line~\lineref{#1}}
\externaldocument[M-]{manuscript}
\externaldocument[S-]{supp-info}
\usepackage[most]{tcolorbox}
\definecolor{block-gray}{gray}{0.9}
\newtcolorbox{myquote}{colback=block-gray,grow to right by=-10mm,grow to left by=-10mm, boxrule=0pt,boxsep=0pt,breakable}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

\renewcommand\thefigure{R\arabic{figure}}

\setcounter{figure}{0}

\doublespacing

We are thankful to the assistant editor James Ross, the senior editor
and David Gibson, and three anonymous reviewers for the insightful
comments to improve the manuscript. In the following sections, you can
find a detailed response to each comment raised during the previous
round of review.

\section{Editor comments}\label{editor-comments}

\begin{quote}
``I agree that your manuscript is much improved and now only in need of
relatively minor revisions. In addition to addressing the AE and
reviewer's comments, please add terms related to invasive species to the
keywords to aid in discoverability of your paper.''
\end{quote}

Many thanks for the encouraging words. We have included `invasive
species' as a keyword. To make space for this keyword we have removed
`controllability' which is already highlighted as part of the title.

\section{Associate Editor's comments}\label{associate-editors-comments}

\begin{quote}
``I think this ms now makes much clearer the potential insights to be
gained by applying control theory to interaction networks. In
particular, the authors have done a nice job of explaining how their
findings reinforce existing ideas in the literature on the stability of
ecological networks while also acknowledging limitations. The reviewers
identified a few remaining issues that should be straightforward to
address. I also noticed a few issues''
\end{quote}

Thank you for the positive feedback. We have addressed all of the
specific issues raised and highlight these in red text in our
resubmitted manuscript.

\begin{quote}
``\emph{L9}: I think here and throughout much of the text, species
should be possessive when followed by a noun (''species role" should be
``species' role''). The possessive is currently used in some instances
but is inconsistently applied throughout the ms."
\end{quote}

\emph{Accepted and changed}. We have checked through the manuscript and
corrected this where relevant. Besides \emph{specie's role}, we also
changed instances of \emph{species' contribution}, \emph{species'
degree}, \emph{species' control capacity}, among others.

\begin{quote}
``\emph{L25}: `every invasive species in our dataset was part of it'
Here it was unclear what `it' refers to (the set of critical species),
and the reference to a singular dataset when you have 20 networks in
total was a bit confusing to me.''
\end{quote}

\emph{Accepted and changed}. We have made this clearer as follows: ``and
included every invasive species in our dataset''
(\lr{M-R2-our-dataset}).

\begin{quote}
``\emph{L23-26}: in these lines, you refer to critical species, but I
think the distinction between having high control capacity versus being
critical needs to be defined (briefly) earlier (it is stated in
\emph{L18-19} that you will quantify both control capacity and identify
critical species, but it is not explained how they are distinct).''
\end{quote}

\emph{Accepted and changed}. We now identify the distinction between
high control capacity and being critical on
\lr{M-R2-define-control-capacity}. The text now reads ``It also allows
us to quantify a species control capacity-its relative importance-and
identify the set of species that, collectively, are critical in this
context because they have largest possible control capacity''.

\begin{quote}
``\emph{L27}: unclear what `its' refers to.''
\end{quote}

\emph{Accepted and changed}. We now explicitly state ``species' degree''
in \lr{M-R2-species-degree}.

\begin{quote}
``\emph{L234, 235}: awkward to refer to species as `they' and
`their'\,''
\end{quote}

\emph{Accepted and changed}. We have changed this to ``it'' and ``its''.

\begin{quote}
``\emph{L296}: `More details in Supporting Information.' is a sentence
fragment.''
\end{quote}

\emph{Accepted and changed}. We have corrected this on
\lr{M-R2-sentence-fragment}. The text now reads ``See Supporting
Information Section \ref{S-visitation-as-proxy} for more details''.

\begin{quote}
``\emph{L343}: I am not sure about citing Table 2 for the first time in
the Discussion. Might it be more useful to cite this in the
Introduction?''
\end{quote}

\emph{Accepted and changed}. We now introduce the glossary at the end of
the introduction on \lr{M-R2-ref-glossary}.

\begin{quote}
``\emph{L111, 114}: `are comprised of' should be `comprised'.
\emph{L339, 432}: `at' should be `in'. \emph{L341}: `allow' should be
`allowed'. \emph{L355}: missing period.''
\end{quote}

\emph{Accepted and changed}. Thank you for pointing out these mistakes,
we corrected all of them.

\section{Reviewer comments \#1}\label{reviewer-comments-1}

\begin{quote}
``I'd like to commend the authors to this revised version of the
manuscript on `Keystoneness, centrality, and the structural
controllability of ecological networks'. I thoroughly enjoyed the
re-written manuscript, which is now considerably more accessible and
clearly targets the community of network ecologists, while not ignoring
the importance of the applied relevance of the findings. Overall, I
don't have any major comments. I guess the language is in parts on the
casual side, yet without compromising the preciseness or clarity of the
text. I noticed a few typos or style errors that the authors may want to
address. Congratulations on this thought-provoking methodological
synthesis.''
\end{quote}

We are pleased that the reviewer enjoyed the re-written manuscript. We
did too!. We also grateful for the typos and style errors identified, we
have striven to correct them all.

\begin{quote}
``A few references require editing to journal style (e.g.~remove
initials for many citations)''
\end{quote}

\emph{Accepted and changed}. We have corrected all citations. One of
them, however, still shows an author's first name initial, as it is
necessary to distinguish two authors with the same last name
(\lr{M-R2-citation-with-name}).

\begin{quote}
``\emph{L388}: remove repeated `other'. I suggest that the authors
re-write this sentence to improve clarity.''
\end{quote}

\emph{Accepted and changed}. We have removed the second `other'. As
suggested, we have rewritten the sentence to improve clarity
(\lr{M-R2-structural-controllability}). It now reads as follows:
``First, structural controllability emphasizes that the effect a species
has on the abundance of other species is not independent of the effects
of those other species in their community''.

\begin{quote}
``Reference list needs updating, e.g.~Ne'eman et al. is not 2009 but
2010.''
\end{quote}

\emph{Accepted and changed}. Thanks for pointing this error out. We've
corrected the reference.

\begin{quote}
``\emph{L314}: open bracket missing. \emph{L358}: change `instead' to
`Instead'. \emph{L381}: remove second comma. \emph{Fig. S6} header of
plot b). correct spelling of `critical'.''
\end{quote}

\emph{Accepted and changed}. Thank you for pointing this out, made all
these changes as indicated.

\section{Reviewer comments \#2}\label{reviewer-comments-2}

\begin{quote}
``I think the concepts presented in this manuscript are a useful
addition to the literature on management of ecological networks. I agree
with the authors that it is important that the discussion surrounding
what constitutes a `keystone' species should go beyond centrality
metrics. I found the results and concepts to be generally well
presented, particularly graphically, and the authors present a thorough
sensitivity analysis to various assumptions.
\end{quote}

We are very thankful for the positive feedback but also for the issues
highlighted. We have aimed to address them all.

\begin{quote}
``The methods are mostly clear but there is no description of how the
`likelihood of being a superior node' is calculated.''
\end{quote}

\emph{Accepted and changed}. Thanks for pointing this out. We have
modified the sentence mentioning `the likelihood of being a superior
node' in the first paragraph of the section \ref{S-reciprocal-links} to
``\ldots{}and whether it is a superior node\ldots{}'' to avoid
confusion. Then we define the `probability of being a superior node' at
the end of the section: ``It is important to note that a species that is
found to be a superior node under a particular control configuration,
will be so in all possible control configurations for the particular
non-reciprocal network. We, therefore denote the average of \(\sigma\)
across all non-reciprocal versions as the probability of the species
being a superior node in its community''.

\begin{quote}
``I would suggest including clarification on the limitations of using
control capacity to summarise species' importance, and examples to help
the reader build an intuition regarding the control capacity metric. As
pointed out, networks that give rise to a species having a high control
capacity can be quite distinct to networks that give rise to a species
having a high value of particular centrality metrics. Providing an
example in the Discussion of, for instance, those species that have a
control capacity of 1 but a low degree would be very useful.''
\end{quote}

\emph{Need Daniel's input}

\begin{quote}
``Jargon occasionally obfuscates the concepts being introduced and there
are places where I thought the language used was vague (esp. if terms
being used have particular meanings depending on the field you're coming
from). I would suggest clarifying, or rewording, the following: `desired
control trajectory' - does this mean `final state'?''"
\end{quote}

\emph{Accepted and changed}: We have attempted to clarify this sentence
as follows: Finding out exactly how difficult it is to control a network
depends strongly on the particularities of the desired control
trajectory (i.e.~the path to the desired final state) as well as the
dynamical relationship between nodes.
\lr{M-R2-desired-control-trajectory}.

\begin{quote}
``\emph{L1364}: `dynamic persistence'\,''
\end{quote}

\emph{Accepted and changed}: Thank you. To avoid confusion we have
removed the word `dynamic' and now the sentence reads ``\ldots{}our
results suggest that structural controllability is closely related to
the persistence\ldots{}'' (\lr{M-R2-dynamic-persistence}). It conserves
its meaning without losing much precision.

\begin{quote}
``\emph{L1370}: `dynamic stability'\,''
\end{quote}

\emph{Accepted and changed}: Contrastingly, in this case the word
dynamic distinguishes a particular facet of stability. We mean the
ability of the system to return to an equilibrium state after a
perturbation. We now include this definition on
\lr{M-R2-dynamic-stability} to make the concept clearer.

\begin{quote}
``\emph{P9}: `ecological objective' - simply `objective'?''
\end{quote}

\emph{Accepted and changed}: We have changed this to simply `objective',
as suggested (\lr{M-R2-ecol-objective}).

\begin{quote}
``\emph{L405}: `central in their community'\,''
\end{quote}

\emph{Accepted and changed}: We've changed the wording to avoid
ambiguity. Now the text reads ``\ldots{}to attain high centrality in
their communities\ldots{}'' (\lr{M-R2-central-in-their-communities}).

\begin{quote}
``\emph{L406}: 'inherently different. The sentence spanning lines
407-410 is quite confusing - please clarify.
\end{quote}

\emph{Accepted and changed}: We rephrased this line as follows: ``We
found that it is not that invasive plants have some different mechanism
for influencing the community compared to their native counterparts''
\lr{M-R2-inherrently-different}.

\begin{quote}
``\emph{L430}: `marginally affected by the topology of their
interactions'\,''
\end{quote}

\emph{Accepted and changed}: We have rephrased this line as: ``By
accounting for network dynamics (even if in a simple way), structural
stability incorporates more ecological realism, especially in the
extreme scenario in which the structure of interactions within the
community only marginally affects the community's state. Both native
species and invasive plants tended to attain a high control capacity if
they were important to network persistence, were abundant, and depended
little on other species'' (\lr{M-R2-marginally-affected}).

\begin{quote}
``\emph{L436} and \emph{L441}: `dominate the state {[}of their
community{]}'\,''
\end{quote}

\emph{Need Daniel's input}

\begin{quote}
``\emph{Discussion}: `species' strength'''
\end{quote}

\emph{Accepted and changed}: In the Methods section we deine species'
strength as the sum of a species' visits (\lr{M-R2-species-strength-a}).
For brevity, we did not define it again in the Discussion, but now we
have improved the text and briefly describe it when we mention it there
too. The text now reads ``When controlling for a species' visitation
strength (the sum of a species' visits), which is indirectly a proxy of
its abundance\ldots{}'' (\lr{M-R2-species-strength-b}).

\begin{quote}
``Minor typos: \emph{Figure S11} caption: `likelyhood' -\textgreater{}
likelihood. \emph{L388}: `other other'. `Driver node density' (supp p12)
-\textgreater{} `size of the minimum driver node set'?''
\end{quote}

\emph{Accepted and changed}: All typos have been corrected as indicated.

\begin{quote}
``\emph{Figure S1b} is referred to (supp. p4 ) but doesn't exist in the
figure''
\end{quote}

\emph{Accepted and changed}: Thanks for noticing this issue. We now
refer to the figure as ``Figure S1 (center panel)''.

\begin{quote}
``\emph{References}: several references use the term `plantPollinator',
or `\ldots{}'\,''
\end{quote}

\emph{Accepted and changed}: We have corrected the reference that
included `plantPollinator' in the title. We have left the references
that include an ellipsis in the author list as they are, as the Journal
of Ecology Author Guidelines indicate that this is the preferred format
when the reference has more than seven authors.

\section{Reviewer comments \#3}\label{reviewer-comments-3}

\begin{quote}
``In the literature on ecological networks, the term keystone species
has often been tossed around without any clear definition. I was pleased
to read that here the authors have developed a cleaner, more effective
way to identify keystone species in ecological networks than methods
that have been proposed before. I think the authors have made the
correct choice in emphasizing the conceptual aspects of the approach,
and do so quite well. The background on control theory was complex but
manageable to read and understand in this context.''
\end{quote}

We are very thankful for the positive feedback.

\begin{quote}
``Overall I found the paper well-written and sound, but there were a few
questions I had that could improve clarity in the methods used. In the
glm that used nD as the response variable, is there any concern that
including N (number of species) as a predictor may cause a problem when
the response is D/N? I can understand why this was included, but I can
also imagine this resulting in models that show an effect of N when
there is none (I could be wrong on this).''
\end{quote}

\emph{Need Daniel's input}

\begin{quote}
``In the species level glm, it was not entirely clear if models were fit
to each network individually or if all species were lumped together (the
former is what I think was done based on my reading of the text).
Regardless I wonder if it would change the results if a mixed effects
model was fit to lumped data, using network as a random effect.''
\end{quote}

### 2019-01-13 paper/drafts_and_revisions/2019-01-14/manuscript.tex
Add revision with most responses to reviewers

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  An important dimension of a species role is its ability to alter the
  state and maintain the diversity of its community. Centrality metrics
  have often been used to identify these species, which are sometimes
  referred as ``keystone'' species. However, the relationship between
  centrality and keystoneness is largely phenomenological and based
  mostly on our intuition regarding what constitutes an important
  species. While centrality is useful when predicting which species
  extinctions could cause the largest change in a community, it says
  little about how these species could be used to attain or preserve a
  particular community state.
\item
  Here we introduce structural controllability, an approach that allows
  us to quantify the extent to which network topology can be harnessed
  to achieve a desired state. It also allows us to quantify a species
  control capacity---its relative importance---and identify the set of
  species that, collectively, are critical in this context. We
  illustrate the application of structural controllability with ten
  pairs of uninvaded and invaded plant-pollinator communities.
\item
  We found that the controllability of a community is not dependent on
  its invasion status, but on the asymmetric nature of its mutual
  dependences. While central species were also likely to have a large
  control capacity, centrality fails to identify species that, despite
  being less connected, were critical in their communities.
  Interestingly, this set of critical species was mostly composed of
  plants and every invasive species in our dataset was part of it. We
  also found that species with high control capacity, and in particular
  critical species, contribute the most to the stable coexistence of
  their community. This result was true, even when controlling for its
  degree, abundance/interaction strength, and the relative dependence of
  their partners.
\item
  \emph{Synthesis}: Structural controllability is strongly related to
  the stability of a network and measures the difficulty of managing an
  ecological community. It also identifies species that are critical to
  sustain biodiversity and to change or maintain the state of their
  community and are therefore likely to be very relevant for management
  and conservation.
\end{enumerate}

\textbf{Keywords:} Keystone species, management interventions,
mutualism, network control theory, plant population and community
dynamics, species importance, control capacity, structural stability,
controllability

@@

A major goal in ecology is to understand the roles played by different
species in the biotic environment. Within community ecology, a
complex-systems approach has led to the development of a variety of
analytical and simulation tools with which to compare and contrast the
roles of species embedded in a network of interactions (J. Bascompte \&
Stouffer, 2009; Coux, Rader, Bartomeus, \& Tylianakis, 2016; Guimerà \&
Amaral, 2005; Stouffer, Sales-Pardo, Sirer, \& Bascompte, 2012). A
particularly relevant dimension of any species' role is its ability to
alter the abundance of other species and the state of the
community---since changes of this nature can have knock-on effects on
ecosystem function, diversity, processes, and services (Thompson et al.,
2012; Tylianakis, Didham, Bascompte, \& Wardle, 2008; Tylianakis,
Laliberté, Nielsen, \& Bascompte, 2010). This ability is sometimes
referred to as a species' ``keystoneness'' (Mills \& Doak, 1993).

A significant proportion of the network tools used to estimate species
roles in this context rely on the calculation of a species'
centrality---a relative ranking of its positional importance that
originally stems from social-network research (Friedkin, 1991; Martín
González, Dalsgaard, \& Olesen, 2010). Generally speaking, central
species tend to be better connected and consequently are more likely to
participate in the network's ``food chains''. Because species that
participate in more chains are more likely to affect the abundances of
other species, centrality metrics have often been used to identify
keystone species in the community (Jordán, Benedek, \& Podani, 2007).
Centrality metrics have been shown to be useful tools to rank species in
regard to their potential to alter the abundances of other species, in
particular when estimating the probability of secondary extinctions that
may follow the loss of a species (Dunne, Williams, \& Martinez, 2002;
Kaiser-Bunbury, Muff, Memmott, Müller, \& Caflisch, 2010).

@@

Species abundances---and consequently the state of the community as a
whole---are influenced both by the structure of their interactions and
the dynamics of these interactions, including the mechanisms of
self-regulation (Lever, van Nes, Scheffer, \& Bascompte, 2014). However,
community and population dynamics can be modelled in innumerable ways,
and empirical support for one versus another is often still ambiguous
(Holland, DeAngelis, \& Bronstein, 2002). The alternative approach
should, therefore, ideally acknowledge ecosystem dynamics, but without
being overly dependent on the particular choices of how they are
characterised. Among the various possibilities \emph{structural
controllability}, a branch of control theory, appears to be a strong
candidate (Isbell \& Loreau, 2013). Control theory is a widely-studied
branch of engineering used to determine and supervise the behaviour of
dynamical systems (A. E. Motter, 2015). It is inherently designed to
deal with system feedbacks and its application has recently been
expanded to complex networks (Lin, 1974; Liu \& Barabási, 2016).
Consistent with long-standing ecological questions, advances in
structural controllability have established a clear link between the
structure of the network and the way nodes affect each other. Unlike
centrality indices, however, this link is not based on a priori
assumptions between network metrics and keystoneness but is instead
based on well-established advances in both dynamical and complex-systems
theory (A. E. Motter, 2015).

@@

Here, we apply methods from structural controllability to a particular
ecological problem and show how it can be used to generate insight into
the role of species in an ecological network. Specifically, we outline
the approach using a set of ten pairs of uninvaded and invaded
plant-pollinator communities. We use invaded communities because there
is strong empirical evidence showing that invasive species play an
important role shaping the abundances of other species, something which
is particularly true in these ten networks (Bartomeus, Vilà, \&
Santamaría, 2008; Lopezaraiza\textendash{}Mikel, Hayes, Whalley, \&
Memmott, 2007). This choice thus offers us an opportunity to explicitly
contrast our theoretical observations with empirical evidence. Moreover,
empirical observations indicate that steering the state of some
communities---for example during ecosystem restoration or invasive
species removal---can be a very difficult task (Woodford et al., 2016).
Therefore, we first ask whether there are differences between the
controllability of invaded and uninvaded networks. We then expand
existing methods from control theory to effectively link the
\emph{controllability} of a network with the role of its constituent
species. We ask---from a control-theoretic perspective---whether there
are key differences between species in the role they play at driving the
state of the community and explore the ecological factors related to
these differences. This allows us to identify species that might be
critical for network control and show that they have a larger than
expected impact on the stable coexistence of the community. Finally, we
compare the proposed approach to current methods based on species
centrality and show how these methods are indeed valuable but ultimately
paint a limited picture in regard to the ``keystoneness'' of a species.

@@

We used ten paired pollination communities to apply the
control-theoretic approach. Each community pair was composed of a
community invaded by a plant and a community free of the invasive
species. Four pairs correspond to natural or semi-natural vegetation
communities in the city of Bristol, UK (Lopezaraiza\textendash{}Mikel et
al., 2007). These communities are comprised of 19--87 species (mean 55),
and non-invaded plots were obtained by experimentally removing all the
flowers of the invasive species \emph{Impatients grandulifera}. The
other six pairs were obtained from lower diversity Mediterranean
shrublands in Cap de Creus National Park, Spain (Bartomeus et al.,
2008). These communities are comprised of 30--57 species (mean 38); in
contrast to the above, uninvaded communities were obtained from plots
that had not yet been colonised by either of the invasive species
\emph{Carpobrotus affine acinaciformis} or \emph{Opuntia stricta}. The
structure of all these communities was defined by the pollinator
visitation frequency, which has been shown to be an appropriate
surrogate for interspecific effects in pollination networks (J.
Bascompte, Jordano, \& Olesen, 2006; Diego P. Vázquez, Morris, \&
Jordano, 2005). Full details about the empirical networks can be found
in the Supporting Information Section \ref{S-empirical-networks}.

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  An important dimension of a species' role is its ability to alter the
  state and maintain the diversity of its community. Centrality metrics
  have often been used to identify these species, which are sometimes
  referred as ``keystone'' species. However, the relationship between
  centrality and keystoneness is largely phenomenological and based
  mostly on our intuition regarding what constitutes an important
  species. While centrality is useful when predicting which species'
  extinctions could cause the largest change in a community, it says
  little about how these species could be used to attain or preserve a
  particular community state.
\item
  Here we introduce structural controllability, an approach that allows
  us to quantify the extent to which network topology can be harnessed
  to achieve a desired state. \R{R2-define-control-capacity} It also
  allows us to quantify a species' control capacity---its relative
  importance---and identify the set of species that, collectively, are
  critical in this context because they have largest possible control
  capacity. We illustrate the application of structural controllability
  with ten pairs of uninvaded and invaded plant-pollinator communities.
\item
  We found that the controllability of a community is not dependent on
  its invasion status, but on the asymmetric nature of its mutual
  dependences. While central species were also likely to have a large
  control capacity, centrality fails to identify species that, despite
  being less connected, were critical in their communities.
  \R{R2-our-dataset} Interestingly, this set of critical species was
  mostly composed of plants and included every invasive species in our
  dataset. We also found that species with high control capacity, and in
  particular critical species, contribute the most to the stable
  coexistence of their community. \R{R2-species-degree} This result was
  true, even when controlling for the species' degree,
  abundance/interaction strength, and the relative dependence of their
  partners.
\item
  \emph{Synthesis}: Structural controllability is strongly related to
  the stability of a network and measures the difficulty of managing an
  ecological community. It also identifies species that are critical to
  sustain biodiversity and to change or maintain the state of their
  community and are therefore likely to be very relevant for management
  and conservation.
\end{enumerate}

\textbf{Keywords:} Invasive species, keystone species, management
interventions, mutualism, network control theory, plant population and
community dynamics, species' importance, control capacity, structural
stability

@@

A major goal in ecology is to understand the roles played by different
species in the biotic environment. Within community ecology, a
complex-systems approach has led to the development of a variety of
analytical and simulation tools with which to compare and contrast the
roles of species embedded in a network of interactions (Bascompte \&
Stouffer, 2009; Coux, Rader, Bartomeus, \& Tylianakis, 2016; Guimerà \&
Amaral, 2005; Stouffer, Sales-Pardo, Sirer, \& Bascompte, 2012). A
particularly relevant dimension of any species' role is its ability to
alter the abundance of other species and the state of the
community---since changes of this nature can have knock-on effects on
ecosystem function, diversity, processes, and services (Thompson et al.,
2012; Tylianakis, Didham, Bascompte, \& Wardle, 2008; Tylianakis,
Laliberté, Nielsen, \& Bascompte, 2010). This ability is sometimes
referred to as a species' ``keystoneness'' (Mills \& Doak, 1993).

A significant proportion of the network tools used to estimate species'
roles in this context rely on the calculation of a species'
centrality---a relative ranking of its positional importance that
originally stems from social-network research (Friedkin, 1991; Martín
González, Dalsgaard, \& Olesen, 2010). Generally speaking, central
species tend to be better connected and consequently are more likely to
participate in the network's ``food chains''. Because species that
participate in more chains are more likely to affect the abundances of
other species, centrality metrics have often been used to identify
keystone species in the community (Jordán, Benedek, \& Podani, 2007).
Centrality metrics have been shown to be useful tools to rank species in
regard to their potential to alter the abundances of other species, in
particular when estimating the probability of secondary extinctions that
may follow the loss of a species (Dunne, Williams, \& Martinez, 2002;
Kaiser-Bunbury, Muff, Memmott, Müller, \& Caflisch, 2010).

@@

Species' abundances---and consequently the state of the community as a
whole---are influenced both by the structure of their interactions and
the dynamics of these interactions, including the mechanisms of
self-regulation (Lever, van Nes, Scheffer, \& Bascompte, 2014). However,
community and population dynamics can be modelled in innumerable ways,
and empirical support for one versus another is often still ambiguous
(Holland, DeAngelis, \& Bronstein, 2002). The alternative approach
should, therefore, ideally acknowledge ecosystem dynamics, but without
being overly dependent on the particular choices of how they are
characterised. Among the various possibilities \emph{structural
controllability}, a branch of control theory, appears to be a strong
candidate (Isbell \& Loreau, 2013). Control theory is a widely-studied
branch of engineering used to determine and supervise the behaviour of
dynamical systems (Motter, 2015). It is inherently designed to deal with
system feedbacks and its application has recently been expanded to
complex networks (Lin, 1974; Liu \& Barabási, 2016). Consistent with
long-standing ecological questions, advances in structural
controllability have established a clear link between the structure of
the network and the way nodes affect each other. Unlike centrality
indices, however, this link is not based on a priori assumptions between
network metrics and keystoneness but is instead based on
well-established advances in both dynamical and complex-systems theory
(Motter, 2015).

@@

Here, we apply methods from structural controllability to a particular
ecological problem and show how it can be used to generate insight into
the role of species in an ecological network. Specifically, we outline
the approach using a set of ten pairs of uninvaded and invaded
plant-pollinator communities. We use invaded communities because there
is strong empirical evidence showing that invasive species play an
important role shaping the abundances of other species, something which
is particularly true in these ten networks (Bartomeus, Vilà, \&
Santamaría, 2008; Lopezaraiza\textendash{}Mikel, Hayes, Whalley, \&
Memmott, 2007). This choice thus offers us an opportunity to explicitly
contrast our theoretical observations with empirical evidence. Moreover,
empirical observations indicate that steering the state of some
communities---for example during ecosystem restoration or invasive
species removal---can be a very difficult task (Woodford et al., 2016).
Therefore, we first ask whether there are differences between the
controllability of invaded and uninvaded networks. \R{R2-ref-glossary}
We then expand existing methods from control theory to effectively link
the \emph{controllability} (Table \ref{tab:tab-glossary}) of a network
with the role of its constituent species. We ask---from a
control-theoretic perspective---whether there are key differences
between species in the role they play at driving the state of the
community and explore the ecological factors related to these
differences. This allows us to identify species that might be critical
for network control and show that they have a larger than expected
impact on the stable coexistence of the community. Finally, we compare
the proposed approach to current methods based on species' centrality
and show how these methods are indeed valuable but ultimately paint a
limited picture in regard to the ``keystoneness'' of a species.

@@

We used ten paired pollination communities to apply the
control-theoretic approach. Each community pair was composed of a
community invaded by a plant and a community free of the invasive
species. Four pairs correspond to natural or semi-natural vegetation
communities in the city of Bristol, UK (Lopezaraiza\textendash{}Mikel et
al., 2007). These communities comprised 19--87 species (mean 55), and
non-invaded plots were obtained by experimentally removing all the
flowers of the invasive species \emph{Impatients grandulifera}. The
other six pairs were obtained from lower diversity Mediterranean
shrublands in Cap de Creus National Park, Spain (Bartomeus et al.,
2008). These communities comprised 30--57 species (mean 38); in contrast
to the above, uninvaded communities were obtained from plots that had
not yet been colonised by either of the invasive species
\emph{Carpobrotus affine acinaciformis} or \emph{Opuntia stricta}. The
structure of all these communities was defined by the pollinator
visitation frequency, which has been shown to be an appropriate
surrogate for interspecific effects in pollination networks (Bascompte,
Jordano, \& Olesen, 2006; Vázquez, Morris, \& Jordano, 2005). Full
details about the empirical networks can be found in the Supporting
Information Section \ref{S-empirical-networks}.

The first step in applying methods of control theory is to construct a
directed network that is able to provide an indication of the extent to
which species affect each other's abundance. In some ecological
networks, establishing the directionality can be relatively
straightforward, for example when links represent biomass transfer or
energy flow (Isbell \& Loreau, 2013). In pollination networks, however,
this directionality is less obvious as both species can, in theory,
benefit from the interaction. We overcome that obstacle by noting that
the extent to which species \emph{i} affects species \emph{j} relative
to the extent to which \emph{j} affects \emph{i} can be summarised by
their interaction asymmetry (Bascompte et al., 2006). This asymmetry is
given by
\[a(i,j) = a(j,i) = \frac{\left | d_{ij}-d_{ji} \right |}{max\left ( d_{ij}, d_{ji} \right )},\]
where the dependence of plant \emph{i} on pollinator \emph{j},
\emph{d\textsubscript{ij}}, is the proportion of the visits from
pollinator \emph{j} compared to all pollinator visits to plant \emph{i}.
Previous research has shown that mutualistic interactions are often
highly asymmetric in natural communities; in other words, if a plant
species is largely dependent on a pollinator species, that pollinator
tends to depend rather weakly on the plant (and vice versa). We
therefore create a directed link from species \emph{i} to species
\emph{j} when \(d_{ij}-d_{ji} \geq 0\) to establish the most likely
direction of control between a species pair (Figure
\ref{fig:fig-example-networks}a). Sometimes there is no observed
asymmetry between species pairs (\(d_{ij}=d_{ji}\)), and we cannot infer
a dominant direction of control. When this occurs, we deem both species
to be equally likely to affect each other and leave a reciprocal
interaction between them (a link from \emph{i} to \emph{j} and another
from \emph{j} to \emph{i}). By basing the direction of the links on the
asymmetry of their dependence, we are able to generate a network that is
consistent with the dynamics of the community while satisfying the
requirements of structural controllability. This allows us to calculate
the controllability of the networks and investigate whether there are
differences between invaded and uninvaded communities.
